Characterizing subarctic biomes for land surface modeling of pollution and climate risk

### 2021-04-19 VegOzone.tex
Improve wording
Abstract merged with Lisa's suggestions

\begin{abstract}
Here we assess the importance of characterization of key meteorological variables and subarctic biomes on the modeled susceptibility of subarctic vegetation to ozone used in pollution and climate risk assessments. Future climate changes will manifest in both direct and indirect effects influencing the species' exposure to air pollution. A likely increase in the frequency of persistent heatwaves in the future will inflict heat-related stress while generating favorable conditions for the formation of ground level ozone, e.g. due to wildfires.\\
We have monitored ground level ozone concentrations at a high latitude site in Northern Norway during the growing season 2018/19. Due to data acquisition problems, ozone concentrations for two weeks in July 2018 were missing from our record. We present a reconstruction based on a probability density function (PDF) and considering the Swedish and Finnish atmospheric monitoring sites. We review available ground level ozone reanalysis products for the region and find only CAMS Regional Air Quality reanalysis ensemble satisfactory.
In terms of anomalies from a multi-annual mean, ozone concentrations were only significantly enhanced in April/May 2018 before the start of the growing season. Temperatures and global irradiance showed consecutively significant positive anomalies during Spring and Summer 2018. In 2019 only early spring and late summer were significantly warmer, while colder during summer. Precipitation was lower in both years but displayed a large variability.\\
We investigate the difference in uptake of ozone using the $\mathrm{DO_3SE}$ model, for typical vegetation at the location. We find that bespoke parameterization for subarctic vegetation is essential for an accurate estimation of Phyto-toxic Ozone Dose (POD) and an assessment of damage risk. Accurate estimates of the start and length of growing season are also crucial (up to $15\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ deviation in $\mathrm{POD_1}$).\\
We find that temperature and its derivative vapor pressure deficit are likely the most important climate variables while the variability of ozone concentration within climatological reasonable limits plays a minor role.
The use of subarctic parameterization suggests an improvement in risk assessment as POD values are increased compared to standard expectations; corroborated by the evidence of visible injury in the Svanhovd ozone garden. We estimate a total biomass reduction in Norway spruce of $2.5-3\,\unit{\%}$, downy birch by $15.5-17.4\,\unit{\%}$ and an above-ground biomass reduction in perennial grassland of $12.4-13.7\,\unit{\%}$.  
\end{abstract}

@@

\subsubsection{Ground-level ozone observations}
\label{subsubsec:ebas}
All ground-level ozone observations used in this study are available from the EBAS database operated by \citet{NILU_EBAS}. In Table~\ref{tab:ebas_obs}, we give an overview of the exact locations and time range of operational data taking. All stations with available long-term observations are located on higher ground than Svanhovd which will result in relatively higher ground-level ozone concentrations in comparison as ozone abundance, in general, decreases with decreasing altitude in the absence of local precursors \citep[e.g.,][]{AB:Klingberg2009}. The periphery vegetation at all sites is similar and consists mainly of pine forests, birch, and heath. However, at higher resolution the vegetation varies substantially. Ozone measurements at Pallas are taken on top of mount Sammaltunturi $100\,\unit{m}$ above treeline (pine forest) and mainly consists of low vascular plants, mosses, and lichen \citep{BER:Hatakka2003}. At Svanhovd, the ozone measurements are taken amidst a meadow surrounded by agriculturally managed herbaceous plants and grass adhering to the WMO standards {\bf TODO: Reference \url{https://library.wmo.int/index.php?id=12407&lvl=notice_display}} (Fig.\ref{fig:svanhovd_research_station}).

@@

For downy birch (Fig.~\ref{fig:pody_rel}a) all simulations exceed the CL by $5-25\,\unit{mmol\,\chem{O_3}\,m^{-2}}$. Downy birch displays the largest variance of ozone uptake among all species or PFTs ($10-31\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). In particular, the sensitivity to the light threshold is especially pronounced, e.g. $\Delta\mathrm{POD_1} = ^{+2.5}_{-3.6}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in case of the boreal type without bespoke GS in 2019. Without bespoke GS, 2018 displays a higher ozone damage risk than 2019. However, taking the later start and overall shorter GS in 2018 into account, this is reversed for the boreal acclimation type. This can be partly explained by very high light response factors (over $f_\mathrm{light}>0.8$) in the beginning of the GS in these simulations. Whether this is a bug or a feature of the $\mathrm{DO_3SE}$ model is not yet clear. SWP induced reduction in ozone uptake is only effective in 2018 and without bespoke GS. The temperature acclimation dependent bias in ozone uptake deduced for bespoke GS ranges between $(1.4-3.6)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.7-7.7)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The maximum systematic deviations deduced from the $-20\,\unit{\%}$ (superscript) and $+20\,\unit{\%}$ (subscript) variation of the extent of stomatal opening at low light intensities amount to $^{+1.6}_{-2.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ and $^{+1.9}_{-2.7}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$, respectively.

@@

\dataavailability{All data is available from public databases or through institutional access. $\mathrm{DO_3SE}$ modeling results can be made available upon request {\bf TODO: Upload to an achieve together with processed, gap filled input data?}.} %% use this section when having only data sets available

@@

\sampleavailability{TEXT} %% use this section when having geoscientific samples available

>>>>

\begin{abstract}
The subarctic region is likely to be at threat from the combined impacts of climate change and ozone pollution due to its unique atmospheric mixing characteristics (which lead to peak ozone concentrations in spring), the influence of the jet stream on regional climate, and its unique vegetation bio-types that are acclimatized to extremes of cold and midnight sun. Unfortunately, to date, only a very few studies have been conducted to assess the response of vegetation to variations in climate and air pollution profiles, at least by comparison with other parts of Europe. This study looks to fill this knowledge gap by examining trends in climate variables and ozone pollution that have occurred over the past few decades. These climatological and ozone data are used to assess the extent to which two recent years (2018 and 2019) deviate from climatic and pollution norms and how these current (and potentially more frequent future) deviations may influence ozone damage to subarctic vegetation. To this end, the quality of available global and regional ozone reanalysis products for Europe is assessed. The analysis suggests that only the CAMS Regional Air Quality (CAMSRAQ) reanalysis ensemble sufficiently reproduces the observed seasonal cycle while systematically underestimating \chem{O_3} concentration in the region. We find that 2018 is an anomalous warm and bright year (in particular in spring and early summer) with higher than average \chem{O_3} concentrations in April/May and a high number of episodic peak \chem{O_3} concentration above $40\,\unit{ppb}$ in June--August, the latter in part attributable to an increase in forest fires in the Northern Hemisphere as well as warmer and sunnier conditions. We apply a flux-metric to determine ozone risk and damage to vegetation as a function of ozone concentration, climate variables, and species-specific physiology. We find that parameterization of this flux-metric for subarctic vegetation bio-types increases estimates of ozone-induced reduction of total biomass by between $2.5$ and $17.4\,\unit{\%}$ suggesting that the UNECE LRTAP assessments that use generic parameterizations would underestimate ozone risk in this region. We conclude that appropriate parameterization for the growing season, and plant physiological responses to light and temperature are particularly important for accurate modeling of seasonal gas exchange and that efforts should be targeted towards accurately defining subarctic bio-type physiological response to climatic variables to improve regional and global scale biogeochemical cycling under current and future climates.
\end{abstract}

@@

\subsubsection{Ground-level ozone observations}
\label{subsubsec:ebas}
All ground-level ozone observations used in this study are available from the EBAS database operated by \citet{NILU_EBAS}. In Table~\ref{tab:ebas_obs}, we give an overview of the exact locations and time range of operational data taking. All stations with available long-term observations are located on higher ground than Svanhovd which will result in relatively higher ground-level ozone concentrations in comparison as ozone abundance, in general, decreases with decreasing altitude in the absence of local precursors \citep[e.g.,][]{AB:Klingberg2009}. The periphery vegetation at all sites is similar and consists mainly of pine forests, birch, and heath. However, at higher resolution the vegetation varies substantially. Ozone measurements at Pallas are taken on top of mount Sammaltunturi $100\,\unit{m}$ above treeline (pine forest) and mainly consists of low vascular plants, mosses, and lichen \citep{BER:Hatakka2003}. At Svanhovd, the ozone measurements are taken amidst a meadow surrounded by agriculturally managed herbaceous plants and grass adhering to the \citet{WMOGuide2018} standards (Fig.\ref{fig:svanhovd_research_station}).

@@

For downy birch (Fig.~\ref{fig:pody_rel}a) all simulations exceed the CL by $5-25\,\unit{mmol\,\chem{O_3}\,m^{-2}}$. Downy birch displays the largest variance of ozone uptake among all species or PFTs ($10-31\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). In particular, the sensitivity to the light threshold is especially pronounced, e.g. $\Delta\mathrm{POD_1} = ^{+2.5}_{-3.6}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in case of the boreal type without bespoke GS in 2019. Without bespoke GS, 2018 displays a higher ozone damage risk than 2019. However, taking the later start and overall shorter GS in 2018 into account, this is reversed for the boreal acclimation type. This can be partly explained by very high light response factors (over $f_\mathrm{light}>0.8$) in the beginning of the GS in these simulations. Whether this is a bug or a feature of the $\mathrm{DO_3SE}$ model is not clear. SWP induced reduction in ozone uptake is only effective in 2018 and without bespoke GS. The temperature acclimation dependent bias in ozone uptake deduced for bespoke GS ranges between $(1.4-3.6)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.7-7.7)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The maximum systematic deviations deduced from the $-20\,\unit{\%}$ (superscript) and $+20\,\unit{\%}$ (subscript) variation of the extent of stomatal opening at low light intensities amount to $^{+1.6}_{-2.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ and $^{+1.9}_{-2.7}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$, respectively.

@@

\dataavailability{All data is available from public databases or through institutional access. $\mathrm{DO_3SE}$ modeling results can be made available upon request {\bf TODO: Upload to an achieve together with processed, gap filled input data on zenodo?}.} %% use this section when having only data sets available

@@

%\sampleavailability{TEXT} %% use this section when having geoscientific samples available
